PubCrawl: A Publication Search Engine

### 2012-05-17 report/paper.tex
Simplify description of PageRank
Don't re-derive the algorithm, just cite the paper and state what we changed

The relationship between academic papers draws a strong analogy with that
between web pages. Just as pages on the web embed links to other pages, papers
cite other papers. This forms a directed graph of paper nodes and citation
edges, where an edge from one paper $A$ to another paper $B$ indicates that $A$
cited $B$.

To exploit the similarity between the two environments, we use a variation of
the PageRank algorithm. PageRank is an algorithm used by the Google search
engine that provides a weighted score for web pages based on the number of
external pages that cite a particular web page, and the PageRank of those
external pages. We use a similar framework for academic papers, where a paper's
score is based on the number of other papers citing it, and the score of those
papers. The most simple form of this captured by the following equation:\\
\begin{equation}
\label{eqn:basic}
S(p)= \Sigma_{c \in C} \frac{S(c)}{D(c)}
\end{equation}
where $p$ is the paper being scored, $C$ is the set of all papers that cite
$p$, $S$ is the scoring function, and $D$ is the degree function (indicating
the number of papers a certain paper cites).

The score of all nodes can be calculated in an iterative manner. All nodes can
be initialized to a default score, such as 1. Then, in each round, Equation
\ref{eqn:basic} can be applied to all nodes. This process is repeated until
node scores converge. However, this scoring function is further complicated by
a couple of features of academic paper relations: cycles and components.

Citation cycles provide one source of complication. In citation cycles, a
paper's chain of citations returns to the paper itself. A number of these
cycles exist in the data set. In general, these cycles act as a score sink,
shifting the score of external nodes into the cycle round after round. Inside
the cycle, the score rotates among nodes, preventing convergence. To reduce the
impact of these cycles, a score source is provided for node each round, as also
done in PageRank. To provide convergence, the total sum of all node scores must
remain constant, so node scores must be normalized per round. This results in
an updated scoring function of:\\
\begin{equation}
\label{eqn:cycle}
S(p)= n(\Sigma_{c \in C} \frac{S(c)}{D(c)}+E(p))
\end{equation}
where $E$ is a source of score, $n$ is the normalization factor, and all other
variables remain as defined.

\begin{figure}[t]
\centering \includegraphics[width=5in]{images/component-graph}
	\caption{Sizes of components in the ACM data set and the number of each
      size's occurrence.}
	\label{comp-size}
\end{figure}

Components in the citation graph are the other source of complication. Papers
in the same component are either both part of the same citation chain or share
a common ancestor or descendant in their respective chain. The ACM data
exhibits 45,571 total components, with one large component of 176,285 papers,
and a large number of smaller components ranging in size from one to ten
papers, as displayed in Figure \ref{comp-size}. Small components are papers
that don't cite many or any other papers in the data set and aren't heavily
cited. This implies that nodes in small components should not be highly ranked.

To adjust for this, nodes are categorized into the components they belong in.
Instead of initializing node scores to be a constant value such as 1, they are
instead initialized to a score linear to the size of the component containing
the nodes. Intuitively, this means that papers in larger components will
receive higher scores, as desired.

A final variation on the basic scoring function is the incorporation of a
damping factor $d$. A user following a citation chain will eventually stop,
possibly at the end of the chain. Intuitively, the farther down a chain a user
traverses, the less likely the papers will relevant and important to the
original query, and the more likely the user will stop. The damping factor
models this behavior by reducing the change produced by the scoring function
as the number of rounds increases. Adapted from PageRank, this results in the
final version of our scoring function:\\
\begin{equation}
\label{eqn:damping}
S(p)= n(\frac{1-d}{N}+d(\Sigma_{c \in C} \frac{S(c)}{D(c)}+E(p)))
\end{equation}
where $d$ is the damping factor, $N$ is the total number of papers/nodes in the
network, and all other variables remain as defined.

>>>>

PageRank is a well-known algorithm for estimating the importance of a web page
as a function of the number of incoming links \cite{brin98, page99}. It is
plausible to extend the algorithm by analogy to academic papers and citations.
However, applying PageRank directly to the ACM citation graph gives suboptimal
results due to the high number of small components. Papers in these components
end with disproportionately high PageRank because further iterations do not
change their initial value very much. Instead of a constant, we set the initial
value of a paper's PageRank to the proportion of all papers contained in its
component.
